\bf Comparing the Power of Plot Designs to Reveal Correlation

### 2015-12-07 paper/Paper_sweave.Rnw
Updated

\newpage
\spacingset{1.45} % DON'T change the spacing!
\newpage
<<setup, echo=FALSE, results=hide, message=FALSE, error=FALSE, warning=FALSE>>=
library(lme4)
library(ggplot2)
library(mnormt)
library(reshape2)
library(nullabor)
library(dplyr)
library(mFilter)
library(grid)
library(gridExtra)

@@

$$p-value = P(X \geq x | H_{0}) = 1 - B_{n, 1/m}(x-1)$$
If all $K$ observers pick the same plot out of a lineup of twenty, we result in a p-value as small as $1/m^K$ . This research used $m= 20$, giving us a significance level (and Type 1 error rate) of $α = 0.05$. The plot of 'true' data is placed randomly amongst 19 null plots to form the lineup.
\begin{figure}[hbtp]
  \centering
<<lineup.example, dependson='setup', fig.width=6, fig.height=6, out.width='0.7\\textwidth', echo=FALSE>>=
set.seed(1234)
a <- gen_true_data(24, 0.95)
td <- melt(a, id="t")
b <- gen_null(24)
b$.n <- NULL
nd_l <- melt(b, id="t")
nd_l$.n <- rep(1:19, rep(24, 19))
pos <- 7
td <- data.frame(.n=rep(pos, nrow(td)), td)
lg <- nd_l$.n < pos
nd_l1 <- data.frame(.n=nd_l$.n[lg], t=nd_l$t[lg], variable=nd_l$variable[lg], 
                    value=nd_l$value[lg])
nd_l2 <- data.frame(.n=nd_l$.n[!lg], t=nd_l$t[!lg], variable=nd_l$variable[!lg], 
                    value=nd_l$value[!lg])
nd_l2$.n <- nd_l2$.n + 1
d <- rbind(nd_l1, td, nd_l2)

@@

$$\hat\pi_{1} - \hat\pi_{2}\pm t_{1-\alpha,2n-1}\sqrt{\hat\pi_{1} (1-\hat\pi_{1}) /n_{1} + \hat\pi_{2} (1-\hat\pi_{2}) /n_{2} }$$

@@

\begin{figure}[hbtp]
  \centering
<<smoothing, dependson='setup', fig.width=6, fig.height=6, out.width='0.7\\textwidth', echo=FALSE>>=
set.seed(123)
uns <- gen_true_data(96, 0.8)
uns_l <- melt(uns, id="t")
p1a <- ggplot(data=uns_l, aes(x=t, y=value, colour=variable)) + geom_line() +labs(title="Unsmoothed Overlaid Lines") +
    theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")
p3a <- ggplot(data=uns, aes(x=X1, y=X2)) + geom_point()+ labs(title="Unsmoothed Scatterplot)") +
    theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")

@@

Each combination of factors was replicated in three different lineups, giving a total of 288 lineups (2 plots x 8 correlations x 6 sample sizes & smoothness combinations). 
The position of the actual data plot in the lineup was randomized. The order that the factors are presented to a subject were also randomized, and each subject saw 5 scatter plots and 5 overlaid lines at a range of all absolute correlation levels. Each subject did not see the same data more than once. Subjects were asked to pick the plot in the lineup that has the strongest association, with basic examples of negatively and positively correlated plots. They also answered with reasons for choosing the plot, and the confidence that they have that this really is the plot showing the strongest correlation. Each subject evaluated 10 lineups, plus a further two trials to eliminate people attempted to game the system. If a trial was not answered correctly, the subject could not evaluate the remaining lineups. If it was answered correctly, the trial data was discarded and the subject moved onto the rest of the lineups. Earlier work [15] has found that repeated evaluations from a subject does not increase their ability to detect the true plot from a lineup, justifying our treatment of the results as independent evaluations. We recruited subjects from Amazon's Mechanical Turk service and ensured that each lineup was viewed by multiple subjects. In total there were 2088 lineup evaluations, however to better control each subject’s individual visual ability we removed data from subjects with less than ten individual evaluations leaving us with 1684 evaluated lineups for the analysis.

\begin{figure}[hbtp]
  \centering
<<designs, dependson='setup', fig.width=6, fig.height=6, out.width='0.7\\textwidth', echo=FALSE>>=
set.seed(101)
a <- gen_true_data(24, 0.95)
b <- gen_true_data(24, -0.95)
ca <- cor(a$X1, a$X2)
cb <- cor(b$X1, b$X2)
a_l <- melt(a, id="t")
b_l <- melt(b, id="t")
p1 <- ggplot(data=a_l, aes(x=t, y=value, colour=variable)) + geom_line() + labs(title="Overlaid line graph") +
      theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")
p2 <- ggplot(data=b_l, aes(x=t, y=value, colour=variable)) + geom_line() + labs(title="Overlaid line graph") +
      theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")
p3 <- ggplot(data=a, aes(x=X1, y=X2)) + geom_point() + labs(title="Scatterplot") +
      theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")
p4 <- ggplot(data=b, aes(x=X1, y=X2)) + geom_point() + labs(title="Scatterplot") +
      theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")
grid.arrange(p1, p2, p3, p4, ncol=2)
@
  \label{designs}
  \caption{The same two data sets in both plot designs. t=24, correlation (left) = 0.946, correlation (right) = -0.943}
\end{figure}

The web site that the Turk workers accessed to complete the task can be seen at http://104.236.245.153:8080/mahbub/turk18/index.html.  You can also try out the tasks, but your results will not be recorded. (The experiment cost $210 to run.)

@@

>>>>

\newpage
\spacingset{1.45} % DON'T change the spacing!
\newpage
<<setup, echo=FALSE, results=hide, message=FALSE, error=FALSE, warning=FALSE>>=
library(lme4)
library(ggplot2)
library(mnormt)
library(reshape2)
library(nullabor)
library(dplyr)
library(mFilter)
library(grid)
library(gridExtra)
library(splitstackshape)
correl <- read.csv("correlations.csv")
full <- read.csv("full.csv") 
Turk <- read.csv("turk.csv")

@@

\begin{equation} \label{eq:1}
  \centering
  p-value = P(X \geq x | H_{0}) = 1 - B_{n, 1/m}(x-1)
\end{equation}

If all $K$ observers pick the same plot out of a lineup of twenty, we result in a p-value as small as $1/m^K$ . This research used $m= 20$, giving us a significance level (and Type 1 error rate) of $α = 0.05$. The plot of 'true' data is placed randomly amongst 19 null plots to form the lineup.
\begin{figure}[htbp]
  \centering
<<lineup.example, dependson='setup', fig.width=6, fig.height=6, out.width='0.7\\textwidth', echo=FALSE>>=
set.seed(1234)
a <- gen_true_data(24, 0.95)
td <- melt(a, id="t")
b <- gen_null(24)
b$.n <- NULL
nd_l <- melt(b, id="t")
nd_l$.n <- rep(1:19, rep(24, 19))
pos <- 7
td <- data.frame(.n=rep(pos, nrow(td)), td)
lg <- nd_l$.n < pos
nd_l1 <- data.frame(.n=nd_l$.n[lg], t=nd_l$t[lg], variable=nd_l$variable[lg], 
                    value=nd_l$value[lg])
nd_l2 <- data.frame(.n=nd_l$.n[!lg], t=nd_l$t[!lg], variable=nd_l$variable[!lg], 
                    value=nd_l$value[!lg])
nd_l2$.n <- nd_l2$.n + 1
d <- rbind(nd_l1, td, nd_l2)

@@

\begin{equation} \label{eq:2}
  \centering
  \hat\pi_{1} - \hat\pi_{2}\pm t_{1-\alpha,2n-1}\sqrt{\hat\pi_{1} (1-\hat\pi_{1}) /n_{1} + \hat\pi_{2} (1-\hat\pi_{2}) /n_{2} }
\end{equation}

@@

\begin{figure}[htbp]
  \centering
<<smoothing, dependson='setup', fig.width=6, fig.height=6, out.width='0.7\\textwidth', echo=FALSE>>=
set.seed(123)
uns <- gen_true_data(96, 0.8)
uns_l <- melt(uns, id="t")
p1a <- ggplot(data=uns_l, aes(x=t, y=value, colour=variable)) + geom_line() +labs(title="Unsmoothed Overlaid Lines") +
    theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")
p3a <- ggplot(data=uns, aes(x=X1, y=X2)) + geom_point()+ labs(title="Unsmoothed Scatterplot)") +
    theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")

@@

Each combination of factors was replicated in three different lineups, giving a total of 288 lineups (2 plots x 8 correlations x 6 sample sizes \& smoothness combinations). 
The position of the actual data plot in the lineup was randomized. The order that the factors are presented to a subject were also randomized, and each subject saw 5 scatter plots and 5 overlaid lines at a range of all absolute correlation levels. Each subject did not see the same data more than once. Subjects were asked to pick the plot in the lineup that has the strongest association, with basic examples of negatively and positively correlated plots. They also answered with reasons for choosing the plot, and the confidence that they have that this really is the plot showing the strongest correlation. Each subject evaluated 10 lineups, plus a further two trials to eliminate people attempted to game the system. If a trial was not answered correctly, the subject could not evaluate the remaining lineups. If it was answered correctly, the trial data was discarded and the subject moved onto the rest of the lineups. Earlier work [15] has found that repeated evaluations from a subject does not increase their ability to detect the true plot from a lineup, justifying our treatment of the results as independent evaluations. We recruited subjects from Amazon's Mechanical Turk service and ensured that each lineup was viewed by multiple subjects. In total there were 2088 lineup evaluations, however to better control each subject’s individual visual ability we removed data from subjects with less than ten individual evaluations leaving us with 1684 evaluated lineups for the analysis.

\begin{figure}[htbp]
  \centering
<<designs, dependson='setup', fig.width=6, fig.height=6, out.width='0.7\\textwidth', echo=FALSE>>=
set.seed(101)
a <- gen_true_data(24, 0.95)
b <- gen_true_data(24, -0.95)
ca <- cor(a$X1, a$X2)
cb <- cor(b$X1, b$X2)
a_l <- melt(a, id="t")
b_l <- melt(b, id="t")
p1 <- ggplot(data=a_l, aes(x=t, y=value, colour=variable)) + geom_line() + labs(title="Overlaid line graph") +
      theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")
p2 <- ggplot(data=b_l, aes(x=t, y=value, colour=variable)) + geom_line() + labs(title="Overlaid line graph") +
      theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")
p3 <- ggplot(data=a, aes(x=X1, y=X2)) + geom_point() + labs(title="Scatterplot") +
      theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")
p4 <- ggplot(data=b, aes(x=X1, y=X2)) + geom_point() + labs(title="Scatterplot") +
      theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")
grid.arrange(p1, p2, p3, p4, ncol=2)
@
  \label{designs}
  \caption{The same two data sets in both plot designs. t=24, correlation (left) = 0.946, correlation (right) = -0.943}
\end{figure}

The web site that the Turk workers accessed to complete the task can be seen at http://104.236.245.153:8080/mahbub/turk18/index.html.  You can also try out the tasks, but your results will not be recorded. (The experiment cost \$210 to run.)

@@

\begin{figure}[htbp]
 \centering
<<results, dependson='setup', fig.width=6, fig.height=8, out.width='0.7\\textwidth', echo=FALSE>>=
Turk <- mutate(Turk, plot = ifelse(test_param =="line", "Overlaid Lines", "Scatterplot"), 
               sign=ifelse(correlation > 0, 1, 0), smoothed=ifelse(smoothed=="yes", 1, 0))
#ggplot(data=subset(Turk, smoothed==0), aes(x=correlation, y=percent, colour=plot)) + geom_smooth() + geom_point() + facet_grid(n~.) + scale_colour_discrete(name = "Plot Design") +  labs(title="Unsmoothed Plot Accuracy", y="Percent Correct", x="True Plot Correlation")
ggplot(data=Turk, aes(x=correlation, y=percent, colour=plot)) + geom_smooth(se=F) + geom_point() + facet_grid(n~smoothed) + 
  scale_colour_discrete(name = "Plot Design") + labs(title="Observed Data", y="Percent Correct", x="Correlation") +
  theme(legend.position="bottom")
@
  \label{results}
  \caption{Response accuracy for unsmoothed data, broken by unsmoothed (left), smoothed (right) and sample size. The dots are data points, the lines represent a naive model average fit. Scatter plots generally showed superior performance, with the overlaid line graphs being particularly weak at revealing negative correlation}

Figure 7 shows the raw results of the survey and Figure 8 shows the difference in power with a $95\%$ confidence interval calculated according to the formula in \ref{eq:2}. Values above zero indicate that the scatter plot had more power to detect the true plot than the overlaid lines.
Performance between the two plot designs appears to be similar for positive correlation, with the scatter plot power gradually improving relative to the overlaid lines as sample size increases. The scatter plot appears to be the superior choice as it retains its power with negative correlation, where the overlaid lines performance suffered. This is not surprising, as positive and negative correlation appear as symmetric graphs in the scatter plot but the two characteristics appear very differently in the overlaid line graph (Figure 6). The right side shows the performance benefits of smoothing overlaid line time series data. Much of the 'noise' in the time series is removed and studying the overall trend shows a greatly improved ability to successfully detect the true data plot when it is negatively correlated. The scatter plot still appears to be the superior plot, but much of the gap between the two is reduced for positive correlation.

\begin{figure}[htbp]
  \centering
<<powerdiff, dependson='results', fig.width=6, fig.height=8, out.width='0.7\\textwidth', echo=FALSE>>=
Turk$correct <- Turk$percent*Turk$attempt
Turk$r <- round(Turk$correlation, 1)
Turk.line <- subset(Turk, test_param=="line")
Turk.scatter <- subset(Turk, test_param=="scatter")
Turk.line <- Turk.line[order(Turk.line$r, Turk.line$n, Turk.line$smoothed),c(10, 13, 16, 19, 20)]
Turk.line <- summarise(group_by(Turk.line, smoothed, n, r), sum(attempt), sum(correct))
Turk.scatter <- Turk.scatter[order(Turk.scatter$r, Turk.scatter$n, Turk.scatter$smoothed),c(10, 13, 16, 19, 20)]
Turk.scatter <- summarise(group_by(Turk.scatter, smoothed, n, r), sum(attempt), sum(correct))
colnames(Turk.scatter) <- c("smoothed", "n", "r", "s.attempt", "s.correct")
colnames(Turk.line) <- c("smoothed", "n", "r", "l.attempt", "l.correct")
prop <- cbind(Turk.line, Turk.scatter[4:5])
prop <- mutate(prop, l.adj = (l.correct+1)/(l.attempt+1), s.adj = (s.correct+1)/(s.attempt+1),
               df = s.attempt + l.attempt - 1, difference=s.adj-l.adj, t = qt(0.975, df),
               upper = difference + t*sqrt(s.adj*(1-s.adj)/s.attempt+l.adj*(1-l.adj)/l.attempt),
               lower = difference - t*sqrt(s.adj*(1-s.adj)/s.attempt+l.adj*(1-l.adj)/l.attempt))
ggplot(data=prop) + geom_line(aes(x=r, y=difference), colour="#99CC99", size=2) + 
                    geom_ribbon(aes(x=r, ymin=lower, ymax=upper), colour="#99CC99", fill="#99CC99", alpha=0.2) + 
                    labs(title="Difference in Power", x="Correlation") + facet_grid(n~smoothed)
@
 \label{powerdiff}
 \caption{Difference in estimated power to reveal the true plot by smoothness and sample size with a 95\% confidence interval. Values above zero indicate the scatter plot is more powerful than the overlaid line graph.}
\end{figure

\begin{figure}[htbp]
  \centering
<<samplesize, dependson='powerdiff', fig.width=6, fig.height=8, out.width='0.7\\textwidth', echo=FALSE>>=
ggplot(data=subset(Turk, smoothed==0), aes(x=correlation, y=percent, colour=factor(n))) + geom_smooth(se=F) + geom_point() + facet_grid(plot~.) + scale_colour_discrete(name = "Sample size") +  labs(title="Effect of sample size, with unsmoothed data", y="Percent Correct", x="Correlation") + theme(legend.position="bottom")}
@
 \label{samplesize}
 \caption{Effects of increasing sample size for unsmoothed data. Both plot designs appear to benefit from increased sample sizes}
\end{figure}

Figure 9 demonstrates the effects of changing sample size for unsmoothed data. The scatter plot has increased power to detect correlation almost universally for any increase. Sample size influences the overlaid lines differently, they appears to benefit for positively correlated true plots, but it does not improve the weak performance in negative correlation.
Figures 10 and 11 give a breakdown of the interactions between the response variables: plot detection, time and confidence. Figure 8 is the histogram of the time taken to record a response, split by plot design and sign of correlation and true plot detection. The time data was highly positively skewed with a maximum of 1760 seconds so the variable was log-transformed. We found that the subject could identify the true plot out of the overlaid line graph 33.6\% of the time and it took on average $e^{3.442}$ = 31.2 seconds to make a choice. Successful detections were slightly faster than unsuccessful decision, taking on average 28.7 seconds to pick the true plot and 32.6 seconds if a null was selected. The scatter plots performed much better, with the true plot picked 52.9\% of the time, with an average of 22.7 seconds required to make a decision. If the true plot was detected it took only 18.6 seconds, and 29.3 seconds for a null plot selection.

\begin{figure}[htbp]
  \centering
<<time, dependson='setup', fig.width=6, fig.height=6, out.width='0.7\\textwidth', echo=FALSE>>=
full <- mutate(full, sign = ifelse(correlation >= 0, 1, 0), n24 = ifelse(n ==24, 1, 0), n48 = ifelse(n ==48,  1, 0), n96 = ifelse(n ==96, 1, 0), plot = ifelse(test_param ==1, "Overlaid Lines", "Scatterplot"))
time <- data.frame(plot=full$plot, time=full$log_time, correct=full$correct, rsgn=full$sign)
#ggplot(data=time, aes(x=time, fill=factor(correct))) +geom_bar(stat="bin") + facet_grid(~plot) +labs(x="Log(Time) Taken", y=element_blank()) + scale_fill_discrete(name = "Correct")
time$rsgn <- factor(time$rsgn, labels=c("r -","r +"))
tm <- summarise(group_by(time, correct, plot, rsgn), m=mean(time))
t1 <- ggplot(data=subset(time, correct==0), aes(x=time)) + geom_histogram(binwidth=0.5) + 
  geom_vline(data=subset(tm, correct==0), aes(xintercept = m), color="red") +
  facet_grid(plot~rsgn) + labs(title="Did not detect", x="Log(Time) Taken", y=element_blank())  
t2 <- ggplot(data=subset(time, correct==1), aes(x=time)) + geom_histogram(binwidth=0.5) + 
  geom_vline(data=subset(tm, correct==1), aes(xintercept = m), color="red") +
  facet_grid(plot~rsgn) + labs(title="Detected", x="Log(Time) Taken", y=element_blank())  
grid.arrange(t1, t2, ncol=2, top = "Log(Time) by plot design, true plot correlation and detection")
@
  \label{time} 
  \caption{Histograms of log time taken to make a decision. r+ and r- indicate the sign of true plot correlation. The vertical red line is the average time of the particular group. On average scatter plots are faster and the true plot is detected more often, with r-.}
\end{figure}

Subjects were asked to give a number from 1 to 5 to rate how confident they were in there choice (1 - most confident, 5 - least confident, average = 3.34). There were no statistically significant differences between plot designs, however subjects were slightly less confident on true plot detections (by $0.25 \pm  0.16, p=0.002$) and every one percent extra time required to make a decision increased confidence (by $0.19 \pm 0.07, p<0.001$). One possible explanation is that subjects often focused on a null plot and picked it confidently. This would suggest that the null plots often produced visually interesting features subjects assumed were caused by the association between plots.

Subjects spending longer on a plot immediately draws two hypotheses:
\begin{enumerate}
  \item The lineup was more difficult and it took longer to determine which plot was best, reducing confidence.
  \item The subject picked a plot relatively easily but spent extra time going over its details and making comparisons to alternative plots, increasing confidence.
\end{enumerate}

The result suggests that the latter hypothesis is more likely to be correct. We also found a negative relationship between plot detection and time spent on a lineup, supporting the major relationship found: quick decisions are less confident but more likely to successfully detect the true plot.

\begin{figure}[htbp]
  \centering
<<confidence, dependson='time', fig.width=6, fig.height=6, out.width='0.7\\textwidth', echo=FALSE>>=
conf <- data.frame(plot=full$plot, conf=full$conf_level, correct=full$correct, rsgn=full$sign)
conf$rsgn <- factor(conf$rsgn, labels=c("r -","r +"))
cm <- summarise(group_by(conf, correct, plot, rsgn), m=mean(conf))
c1 <- ggplot(data=subset(conf, correct==0), aes(x=conf)) + geom_histogram(binwidth=1) + 
  geom_vline(data=subset(cm, correct==0), aes(xintercept = m), color="red") +
  facet_grid(plot~rsgn) + labs(title="Did not detect", x="Confidence", y=element_blank())  
c2 <- ggplot(data=subset(conf, correct==1), aes(x=conf)) + geom_histogram(binwidth=1) + 
  geom_vline(data=subset(cm, correct==1), aes(xintercept = m), color="red") +
  facet_grid(plot~rsgn) + labs(title="Detected", x="Confidence", y=element_blank())  
grid.arrange(c1, c2, ncol=2, top = "Confidence by plot design, true plot correlation and detection")
@
  \label{confidence} 
  \caption{Histograms of reported confidence levels. 1 - Low, 5 - High. Detection had a minor detrimental impact on confidence, whilst time taken increased it. There was no statistically significant difference on confidence level between plot designs.}
\end{figure}

Next we analysed the individual ability of each of our subjects. Figure 12 plots the accuracy of each subject on the y axis versus the average confidence on the x axis. While we expected that subjects with greater visual skills would report lower values of confidence, there does not appear to be any relationship between the two. The accuracy is normally distributed (Jarque-Bera p value = 0.418), with a mean of 42.5\% and a standard deviation of 17.8\%. The strongest subject successfully identified the true plot 91\% of the time, whilst the weakest was never able to identify the true plot. As all subjects were shown a balanced mix of true plot correlations, the extreme difference in individual visual ability is striking. Visual ability will be controlled via a subject specific random effect in the models in the next section.

\begin{figure}[htbp]
  \centering
<<confidence2, dependson='confidence', fig.width=6, fig.height=6, out.width='0.7\\textwidth', echo=FALSE>>=
full$subj_id <- factor(full$ip_address, labels=1:207)
y <- data.frame(table(full$subj_id))
keep2 <- y$Var1[y$Freq>9]
sub <- filter(full, subj_id %in% keep2)
sub <- sub[c(7, 15, 16, 17, 20, 27)]
sub <- summarise(group_by(sub, subj_id), correct = sum(correct), conf=mean(conf_level))
sub <- merge(sub, y, by.x="subj_id", by.y="Var1")
sub <- mutate(sub, prop = correct/Freq)
ggplot(data=sub, aes(x=conf, y= prop)) + geom_point() + labs(title="Accuracy versus Confidence", y = "Accuracy", x = "Average Confidence")
@
  \label{confidence2}
  \caption{Individual subject accuracy versus confidence. There does not appear to be any relationship between the two.}
\end{figure}

There are two major reasons why a subject may fail to detect the true plot out of a lineup. Either the plot design was unable to display correlation for that particular combination of factors (and the analyst would commit a Type II error by not rejecting the null hypothesis), or a null plot that had a greater correlation than the true plot was selected instead (and the analyst would commit a Type I error by rejecting the null hypothesis). We next can analyse these kinds of errors with the pin plots found in Figure 13. The frequency of selection of each plot in each lineup, either true or null, is represented by a pin, placed at the value of the correlation between the two variables shown in that plot. The correlation of the selected plot is on the x-axis and the frequency of selection is plotted on the y axis. The correlation of the simulated (true) data (r=-0.9, -0.7, -0.3, 0.3, 0.7, 0.9) in the plot, the sample size (t=12, 24, 48, 96), and the plot design (unsmoothed/smoothed, scatterplot/lineplot) is used to facet the plots. The blue pins indicate the true plots and the red pins indicate the null plots. If the plot design has the power to display correlation, we would expect the highest correlation plot to be selected, leading to a Type I error where a plot from the null distribution spuriously had high correlation. We find that this is generally not the case. The selections appear to have a wide range of correlations, indicating that if a subject selected a null plot it was typically not due to excess spurious correlation. Instead the failure to select the highest correlated plot indicates that a Type II error was committed rather than Type I as the plot designs did not have the power to differentiate between medium and low correlation plots. However, for the scatter plot, whenever there was a plot with a very strong correlation, usually in the form of the true plot, it was overwhelmingly selected. For the overlaid lines this mostly occurred only for plots with strong positive correlation, which is consistent with its general lack of power to reveal negative relationships in data.

\begin{figure}[htbp]
  \centering
<<pins, dependson='samplesize', fig.width=6, fig.height=10, out.width='0.7\\textwidth', echo=FALSE>>=
correl <- read.csv("correlations.csv")
pin <- full[c(5, 6, 10, 15, 16, 17)]
pin <- cSplit(pin, "response_0", direction="long")
counts <- data.frame(table(pin$pic_id, pin$response_0))
counts$Var1 <- as.integer(counts$Var1)
counts$Var2 <- as.integer(counts$Var2)
colnames(counts) <- c("lineup", "plot", "count")
Turka <- Turk[c(2, 10, 11, 13, 17, 20)]
pin <- merge(counts, Turka, by.x="lineup", by.y="pic_id")
pin$r <- factor(pin$r)
head(pin)
colnames(pin) <- c("lineup", "plot", "count", "smoothed", "true_pos", "n", "design", "r")
pin$base <- 0
pin <- read.csv("pin.csv") #Do this part through R, not excel
pin <- mutate(pin, true = ifelse(plot==true_pos, "True plot", "Null plot"))
pin <- merge(pin, Turk[,c(2, 16)], by.x = "lineup", by.y = "pic_id")
pin <- mutate(pin, percent = count/attempt)
ggplot(data=subset(pin, smoothed==0 & design == "Scatterplot")) + geom_errorbar(aes(x=Correlation, ymin=base, ymax=percent, colour=true)) +
       geom_point(aes(x=Correlation, y=percent, colour=true)) + coord_fixed() + facet_grid(n~r) + labs(title="Unsmoothed Scatterplot", y="Percent Chosen") + theme(legend.position="none")
ggplot(data=subset(pin, smoothed==1 & design == "Scatterplot")) + geom_errorbar(aes(x=Correlation, ymin=base, ymax=percent, colour=true)) +
  geom_point(aes(x=Correlation, y=percent, colour=true)) + coord_fixed()  + facet_grid(n~r) + labs(title="Smoothed Scatterplot", y="Percent Chosen") + theme(legend.position="none")
ggplot(data=subset(pin, smoothed==0 & design == "Overlaid Lines")) + geom_errorbar(aes(x=Correlation, ymin=base, ymax=percent, colour=true)) +
  geom_point(aes(x=Correlation, y=percent, colour=true)) + coord_fixed()  + facet_grid(n~r) + labs(title="Unsmoothed Overlaid Lines", y="Percent Chosen") + theme(legend.position="none")
ggplot(data=subset(pin, smoothed==1 & design == "Overlaid Lines")) + geom_errorbar(aes(x=Correlation, ymin=base, ymax=percent, colour=true)) +
  geom_point(aes(x=Correlation, y=percent, colour=true)) + coord_fixed()  + facet_grid(n~r) + labs(title="Smoothed Overlaid Lines", y="Percent Chosen") + theme(legend.position="none")

#Code to be fixed
@
  \label{pins}
  \caption{Pin plots of the subject selections of each lineup. On the x-axis is the correlation of the plot (there is potentially spurious correlation in the null plots), and on the y-axis is the proportion a plot was chosen. Red pins indicate null plots and blue indicates the true plot, occasionally the true plot does not appear as it has a frequency of zero and is visually blocked by the red pins. The pin plots are split with true plot correlation on the columns and sample size on the rows. Each plot contains information from all three lineups with that unique combination of factors: plot design, smoothness, sample size and true correlation.}
\end{figure}

Occasionally there was an anomaly where a single null plot was selected with a high frequency despite having a weak correlation. Occasionally the null distribution cangenerate visually interesting patterns that draw the analyst’s attention.  We asked subjects for reasons behind their selections and found that choosing on the basis that “the points in the plot were close together”, or had “a unique pattern” more often led to an unsuccessful detection. Conversely, decisions based on “matching the peaks and troughs” of the overlaid lines or “seeing a line formed in the dots” of the scatter plot were more likely to be successful in identification of the true plot. However, these true detections are, on average, made less confidently than unsuccessful evaluations, which indicates that decisions made on the basis of closeness or patterns are more convincing of an association than the more successful reasons.

\begin{figure}[htbp]
  \centering
Insert anomaly lineup 1
  \label{anom1}
  \caption{A lineup of unsmoothed overlaid lines with a sample size of 12 and true correlation of -0.707. The true plot in position 8 was not detected in five evaluations. However, plot 13 was selected four times, (p –value < 0.001) despite having a correlation of only 0.225. When asked for a reason of their selections, subjects answered that the lines were closest together.}
\end{figure}

\begin{figure}[htbp]
  \centering
Insert anomaly lineup 2
  \label{anom2}
  \caption{A lineup of unsmoothed scatter plots with a sample size of 24. The true plot is located in position 4 with a correlation of -0.491, and was selected by one out of eleven observers (p = 0.33, see formula \ref{eq:1}). The plot in position 6 was selected by nine of these eleven observers (p < 0.001) despite having a correlation of -0.037. This null plot shows a unique clustering effect that draws in the observer despite not having any linear relationship in the data, an example of the over-interpretation of randomness. Subjects most often cited the clustering pattern formed as the reason for their selection. The final selection was position 8, with a correlation of -0.226 (p = 0.33).}
\end{figure}

Figures 14 and 15 show lineups where a null plot with a strong pattern led to its selection over the data plot.  Mostly, when a null plot was chosen over the data plot it was because the correlation was stronger, but in a few lineups a null plot with small correlation was selected by many subjects.  The true data plot is in position 4, but 9/11 subjects chose plot 6. Plot 6 is unusual, most of the points are scattered around the vertical median, but there are two outlying points, one with a high y-value and one with a low y-value.  Other concurrent work has suggested that outliers are pre-attentive patterns and this plot selection by the observers is consistent with that. Figure 15 below is another example of where subjects chose a null plot that had smaller correlation than the data plot. Four out of five evaluations selected the null plot in position 13,  instead of the data plot in position 8. Plot 13 has a correlation of 0.225, smaller (in absolute value) than the -0.707 of plot 8.  This selection is in-line with Wertheimer’s gestalt law of common fate. It appears to be strongly correlated in the beginning of the series but breaks apart towards the end. The selection implies that subjects selecting this plot focused on the joint path of the two series early in the plot and ignored the contradictory evidence of no relationship as they diverged later in the plot.  It also confounded by the seeming preference to find positive correlation over negative correlation, and we suspect that if one of the series in plot 8 had the values flipped, then the strong positive correlation would have been noticed.

\section{Modelling Subject Responses}

A mixed effects logit model is used separately for scatter plot and overlaid line evaluations. The tested model is:

$$Power_{i} = \psi\left\{ \beta_{i,0} + \beta_{i,1} \cdot Correlation + \beta_{i,2}\cdot Sign + \beta_{i,3}\cdot Smoothed + \beta_{i,4}\cdot t + \beta_{i,5}\cdot Correlation \cdot Sign + \gamma_{i,1} \cdot subject_{j} \right\}$$

Where $\psi$ is the logit link function, $\psi(x) = \frac{e^{x}}{1+e^{x}}$, $r$ is the true plot correlation, Sign is the direction of the true plot correlation, $i$  is the plot design and $j$ is an observer specific random effect.

Table 1: Mixed Effects Logit Model for the predicted power of plot design. Sign = 1 if correlation > 0, 0 otherwise. Each plot design is estimated separately. Results must be transformed via the logit link function to generate a power prediction. Subjects with less than ten lineup evaluations were removed from the sample. Sample size for scatter plot model: 841, overlaid lines model: 843. The model is fitted using the lme4 package in 
<<powerfit, dependson='setup', echo=FALSE, results=hide, message=FALSE, error=FALSE, warning=FALSE>>=
full$subj_id <- factor(full$ip_address, labels=1:207)

# Need to keep only evaluations done by subject who completed at least 10
x <- data.frame(table(full$subj_id))
keep <- x$Var1[x$Freq>9]
full.sub <- filter(full, subj_id %in% keep)
full.sub$r <- round(full.sub$correlation, 1)
full.sub$rsgn <- ifelse(full.sub$r >= 0, 1, 0)
full.sub$rabs <- abs(full.sub$r)
full.sub$rabs <- factor(full.sub$rabs)
full.sub$plot <- ifelse(full.sub$test_param == 0, "S", "L")
full.sub <- full.sub[,c(1,7, 15,16,17,20:26)]
full.sub$rpos <- ifelse(full.sub$r > 0, full.sub$r, 0)
full.sub$rneg <- ifelse(full.sub$r < 0, full.sub$r, 0)
full.sub$alpha <- log(.05/0.95)

m8l <- glmer(correct~offset(alpha) + r + rsgn + r*rsgn + smoothed + n + (1|subj_id), data=subset(full.sub, plot == "L"), family="binomial")
m8s <- glmer(correct~offset(alpha) + r + rsgn + r*rsgn + smoothed + n + (1|subj_id), data=subset(full.sub, plot == "S"), family="binomial")
@
%Code to make a table to be added
